---
title: Session 8
subtitle: Practical session
date: "2022-06-10"
author:
  - Matt Denwood
theme: metropolis
aspectratio: 169
colortheme: seahorse
header-includes: 
  - \input{../rsc/preamble}
params:
  presentation: TRUE
output:
  beamer_presentation:
      pandoc_args: ["-t", "beamer"]
      slide_level: 2
  html_document: default
---

```{r rendering, eval=FALSE, include=FALSE}
# To render this as PDF (beamer) slides run:
rmarkdown::render('Session_8.Rmd', 'beamer_presentation', params=list(presentation=TRUE))
# And for html:
rmarkdown::render('Session_8.Rmd', 'html_document', params=list(presentation=FALSE))
```

```{r setup, include=FALSE}
source("../rsc/setup.R", local = environment())
```

## Practical session

This is the final session of the course!

We have three options to choose from:

1. Finish up working on practical exercises from previous sessions

2. Work on your own data and ask us for help/advice

3. Look at how to implement custom distributions/functions via a JAGS module

  * See the JAGS_modules subdirectory

## Before you go...

- Feedback on the course would be extremely welcome!
  - Please fill in the stop/start/continue feedback forms provided

. . .

- Remember to keep an eye on the COST action website:
  - http://harmony-net.eu
  - Physical training schools are being run in September and accepting sign-ups now!


```{r include=FALSE}
unlink(cleanup)
```
